---
title: "final_rmd"
output: html_document
names: "Julia Van Goor, Joshua Strauss, Alex Yang"
---
# Set Working Directory
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Owner/Documents/Group1_Bio201_FinalProject/")
```


# Load Packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(phyloseq)
library(vegan)
set.seed(7)
```

#Import Data
```{r}
library(readr)
biographical <- read_delim("~/Group1_Bio201_FinalProject/raw_data/biographical.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
View(biographical)

library(readr)
phylum_processed_df_both_weeks_present_final <- read_delim("~/Group1_Bio201_FinalProject/Phylum Data Frames/phylum_processed_df_both_weeks_present_final.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
View(phylum_processed_df_both_weeks_present_final)

library(readr)
all_data_wkly <- read_delim("~/Group1_Bio201_FinalProject/raw_data/Bigzipfile/all_data_wkly.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
View(all_data_wkly)

```

#Tidy Data
##Butyrate and Propionate df
```{r}
c_all_data_wkly <- all_data_wkly %>%
  rename_all(tolower) %>% 
  select(participant_id, semester, supplement_consumed, frequency, quantity_compliant, study_week, butyrate_mean, propionate_mean) %>% #keep only columns of interest
  filter(study_week == "week1" | study_week == "week3", quantity_compliant == "yes") #curated all data weekly: keep only weeks and supplements of interest

#wide dfs
but_wide <- all_data_wkly %>%
  rename_all(tolower) %>% 
  select(participant_id, semester, supplement_consumed, frequency, quantity_compliant, study_week, butyrate_mean) %>% #keep only columns of interest
  filter(study_week == "week1" | study_week == "week3", quantity_compliant == "yes") %>% #keep only weeks of interest, quantity compliant, and 1x
  spread(study_week, butyrate_mean) %>% 
  mutate(delta_butyrate = week3 - week1) #calculate difference in butyrate 

pro_wide <- all_data_wkly %>%
  rename_all(tolower) %>% 
  select(participant_id, semester, supplement_consumed, frequency, quantity_compliant, study_week, propionate_mean) %>% #keep only columns of interest
  filter(study_week == "week1" | study_week == "week3", quantity_compliant == "yes") %>% #keep only weeks of interest, quantity compliant, and 1x
  spread(study_week, propionate_mean) %>% 
  mutate(delta_propionate = week3 - week1) #calculate difference in propionate

but_pro_wide <- inner_join(x = but_wide, y = pro_wide,
                    by = c("participant_id", 
                           "semester", "supplement_consumed", "frequency")) %>%
  rename(but_wk1 = week1.x, but_wk3 = week3.x, pro_wk1 = week1.y, pro_wk3 = week3.y) %>% 
  select(-starts_with("quantity_compliant")) #all measurements by participant ID
```
##BMI df
```{r}
BMI_data <- biographical %>% 
  rename_all(tolower) %>%
  select(participant_id, sex, age, race_ethnicity, weight_kg, height_meters) %>%
  drop_na(weight_kg, height_meters) %>%
  mutate(BMI = weight_kg/(height_meters)^2) #calculate all BMIs; n = 252

underweight_BMI <- BMI_data %>% 
  filter(BMI < 18.5) #subset by CDC category: underweight - n = 12

healthy_BMI <- BMI_data %>% 
  filter(BMI >= 18.5, BMI < 25) #subset by CDC category: healthy - n = 186

overweight_BMI <- BMI_data %>% 
  filter(BMI >= 25, BMI < 30) #subset by CDC category: overweight - n = 39

obese_BMI <- BMI_data %>% 
  filter(BMI >= 30) #subset by CDC category: obese - n = 15

#total n = 252
```

##Phylum df
```{r}
#Creating Processed Phylum Data Frame:
phylum_processed_df <- phylum_avg_shared %>%
  rename_all(tolower) %>%
  filter(study_week == "week1" | study_week == "week3") %>%
  select("participant_id", "semester", "study_week", "firmicutes", "bacteroidetes <bacteroidetes>") %>%
  rename(., "bacteroidetes" = "bacteroidetes <bacteroidetes>") %>%
  mutate(., f_b_ratio = firmicutes/bacteroidetes)

#Checking Sample Sizes for Consistency by Week:
phylum_processed_df %>%
  filter(study_week == "week1") %>%
  summarise(sample_size = n()) # n = 508

phylum_processed_df %>%
  filter(study_week == "week3") %>%
  summarise(sample_size = n()) # n = 480. 508-480 = 28 inconsistencies. 

#Split by Weeks and Recombine so only samples with both weeks present: 
phylum_week1 <- phylum_processed_df %>%
  filter(study_week == "week1")

phylum_week3 <- phylum_processed_df %>%
  filter(study_week == "week3")
phylum_week3

phylum_processed_df_both_weeks_present <- inner_join(phylum_week1, phylum_week3, by = (c("participant_id")))

#Final Data Frames:
phylum_processed_df
phylum_processed_df_both_weeks_present

phylum_processed_df_both_weeks_present_final <- phylum_processed_df_both_weeks_present %>%
  rename(firmicutes_wk1 = firmicutes.x, bacteroidetes_wk1 = bacteroidetes.x, fb_ratio_wk1 =  
  f_b_ratio.x, firmicutes_wk3 = firmicutes.y, bacteroidetes_wk3 = bacteroidetes.y, fb_ratio_wk3 = 
  f_b_ratio.y, semester = semester.x) %>%
  select(-starts_with("study_week"), -starts_with("semester.y"))
View(phylum_processed_df_both_weeks_present_final)

#Save Plots: 
write_delim(phylum_processed_df, path = "~/Desktop/phylum_processed_df.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
write_delim(phylum_processed_df_both_weeks_present, path = "~/Desktop/phylum_processed_df_both_weeks_present.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
write_delim(phylum_processed_df_both_weeks_present_final, path = "~/Desktop/phylum_processed_df_both_weeks_present_final.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
```

##Joined BMI and scfas
```{r}
BMI_but_pro <- inner_join(x = BMI_data, y = but_pro_wide,
                          by = c("participant_id")) #n = 178

underweight_but_pro <- inner_join(x= underweight_BMI, y = but_pro_wide,
                                  by = c("participant_id")) #n = 10

healthy_but_pro <- inner_join(x= healthy_BMI, y = but_pro_wide,
                                  by = c("participant_id")) #n = 133

overweight_but_pro <-  inner_join(x= overweight_BMI, y = but_pro_wide,
                                  by = c("participant_id")) #n = 25

obese_but_pro <-  inner_join(x= obese_BMI, y = but_pro_wide,
                                  by = c("participant_id")) #n = 10
#total n = 178
```

##Joined BMI, scfas, and phylums
```{r}
BMI_phylum_scfa <- inner_join(x = BMI_but_pro, y = phylum_processed_df_both_weeks_present_final,
                          by = c("participant_id")) #n = 139

underweight_phylum_scfa  <- inner_join(x = underweight_but_pro, y = phylum_processed_df_both_weeks_present_final,
                          by = c("participant_id")) #n = 9

healthy_phylum_scfa <- inner_join(x = healthy_but_pro, y = phylum_processed_df_both_weeks_present_final,
                          by = c("participant_id")) #n = 105

overweight_phylum_scfa <- inner_join(x = overweight_but_pro, y = phylum_processed_df_both_weeks_present_final,
                          by = c("participant_id")) #n = 20

obese_phulum_scfa <- inner_join(x = obese_but_pro, y = phylum_processed_df_both_weeks_present_final,
                          by = c("participant_id")) #n = 5
```

#Analysis
##Broad (All BMIs)
```{r}

```

##Underweight
```{r}

```

##Healthy Weight
###Wk1 F:B and BMI
```{r}
#PLOT
hwk1fb_BMI_plot<-healthy_phylum_scfa %>%
  ggplot(aes(x = fb_ratio_wk1,
             y = BMI)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
 labs(title = "Ratio of Firmicutes to Bacteroidetes by healthy BMI",
         x = "Ratio of Firmicutes to Bacteroidetes",
         y = "BMI")
print(hwk1fb_BMI_plot)

#TEST
hwk1fb_BMI_test <- healthy_phylum_scfa %>%
  lm(fb_ratio_wk1 ~ BMI, data = .) #test relationship
summary(hwk1fb_BMI_test) #view results 
#p-value = 0.3083
#Adjusted R-squared = 0.000465

#Save Plot
save_plot(filename = "figures/hwk1fb_BMI_plot.pdf",
          plot = hwk1fb_BMI_plot,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)
```
The p-value for the linear correlation of the week 1 firmicutes to bacteroidetes ratio and the BMI of "healthy" participants was 0.3083, which is not statistically significant. Furthermore, the R value was an insignificant value close to zero (0.000465). There does not appear to be a statistically significant linear relationship between the week 1 firmicutes to bacteroidetes ratio and BMI among participants whose BMI falls within the CDC's "Healthy" category.

###Wk1 F:B and Wk1 Butyrate
```{r}
#PLOT
hwk1fb_wk1but_plot<-healthy_phylum_scfa %>%
  ggplot(aes(x = fb_ratio_wk1,
             y = but_wk1)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
 labs(title = "Week 1 Ratio of Firmicutes to Bacteroidetes by Butyrate Concentration",
         x = "Ratio of Firmicutes to Bacteroidetes",
         y = "Average Butyrate Concentration (mmol/kg)")
print(hwk1fb_wk1but_plot)

#TEST
hwk1fb_wk1but_test <- healthy_phylum_scfa %>%
  lm(fb_ratio_wk1 ~ but_wk1, data = .) #test relationship
summary(hwk1fb_wk1but_test) #view results 
#p-value = 0.005637
#Adjusted R-squared = 0.06302
```
The p-value for the linear correlation of the week 1 firmicutes to bacteroidetes ratio and the week 1 average butyrate concentration among "healthy" participants was 0.005637, which falls within a 95% confidence interval. However, the R value was an insignificant value close to zero (0.06302). There does not appear to be a statistically significant linear relationship between the week 1 firmicutes to bacteroidetes ratio and the week 1 average butyrate concentration among participants whose BMI falls within the CDC's "Healthy" category.

###Wk1 F:B and Wk1 Propionate
```{r}
#PLOT
hwk1fb_wk1pro_plot<-healthy_phylum_scfa %>%
  ggplot(aes(x = fb_ratio_wk1,
             y = pro_wk1)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
 labs(title = "Week 1 Ratio of Firmicutes to Bacteroidetes by Propionate Concentration",
         x = "Ratio of Firmicutes to Bacteroidetes",
         y = "Average Propionate Concentration (mmol/kg)")
print(hwk1fb_wk1pro_plot)

#TEST
hwk1fb_wk1pro_test <- healthy_phylum_scfa %>%
  lm(fb_ratio_wk1 ~ pro_wk1, data = .) #test relationship
summary(hwk1fb_wk1pro_test) #view results 
#p-value = 0.004754
#Adjusted R-squared = 0.06841
```
The p-value for the linear correlation of the week 1 firmicutes to bacteroidetes ratio and the week 1 average propionate concentration among "healthy" participants was 0.004754, which falls within a 95% confidence interval. However, the R value was an insignificant value close to zero (0.06841). There does not appear to be a statistically significant linear relationship between the week 1 firmicutes to bacteroidetes ratio and the week 1 average propionate concentration among participants whose BMI falls within the CDC's "Healthy" category.

###Wk3 F:B and Wk3 Butyrate
```{r}
#PLOT
hwk3fb_wk3but_plot<-healthy_phylum_scfa %>%
  ggplot(aes(x = fb_ratio_wk3,
             y = but_wk3)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
 labs(title = "Week 3 Ratio of Firmicutes to Bacteroidetes by Butyrate Concentration",
         x = "Ratio of Firmicutes to Bacteroidetes",
         y = "Average Butyrate Concentration (mmol/kg)")
print(hwk3fb_wk3but_plot)

#TEST
hwk3fb_wk3but_test <- healthy_phylum_scfa %>%
  lm(fb_ratio_wk3 ~ but_wk3, data = .) #test relationship
summary(hwk3fb_wk3but_test) #view results 
#p-value = 0.7948
#Adjusted R-squared = -0.009043
```
The p-value for the linear correlation of the week 3 firmicutes to bacteroidetes ratio and the week 3 average butyrate concentration among "healthy" participants was 0.7948, which is not statistically significant. Furthermore, the R value was an insignificant value close to zero (-0.009043). There does not appear to be a statistically significant linear relationship between the week 3 firmicutes to bacteroidetes ratio and the week 3 average butyrate concentration among participants whose BMI falls within the CDC's "Healthy" category.

###Wk3 F:B and Wk3 Propionate
```{r}
#PLOT
hwk3fb_wk3pro_plot<-healthy_phylum_scfa %>%
  ggplot(aes(x = fb_ratio_wk3,
             y = pro_wk3)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
 labs(title = "Week 3 Ratio of Firmicutes to Bacteroidetes by Propionate Concentration",
         x = "Ratio of Firmicutes to Bacteroidetes",
         y = "Average Propionate Concentration (mmol/kg)")
print(hwk3fb_wk3pro_plot)

#TEST
hwk3fb_wk3pro_test <- healthy_phylum_scfa %>%
  lm(fb_ratio_wk3 ~ pro_wk3, data = .) #test relationship
summary(hwk3fb_wk3pro_test) #view results 
#p-value = 0.01675
#Adjusted R-squared = 0.04645
```
The p-value for the linear correlation of the week 3 firmicutes to bacteroidetes ratio and the week 3 average propionate concentration among "healthy" participants was 0.01675, which falls within a 95% confidence interval. However, the R value was an insignificant value close to zero (0.04645). There does not appear to be a statistically significant linear relationship between the week 3 firmicutes to bacteroidetes ratio and the week 3 average propionate concentration among participants whose BMI falls within the CDC's "Healthy" category.

##Overweight
```{r}

```

##Obese
```{r}

```

#Final Conclusions